### 2020-02-07 doc/icfp20/counterpoint.tex
fix checkUnison; more on counterpoint

In this section, we present an implementation of species counterpoint,
based on the formulation given by Fux~\citep{fux-cp}.
The idea is to represent rules for composing ``good'' counterpoint
using dependent types.
[One more sentence here.]

@@

The simplest variant of counterpoint is called the first species
counterpoint.
In first species, we set one note against each note in the cantus
firmus, as in the following example:

[Frog]

In our implementation, we represent each bar as a pitch-interval
pair, and the whole music as a list of such pairs.

\begin{alltt}
PitchInterval : Set
PitchInterval = Pitch \(\times\) Interval
\end{alltt}

First species counterpoint comes with five mandatory rules, 
constraining the beginning, the main body, and the ending of
the music.
Below, we describe these rules one by one, showing the
corresponding Agda code.

@@

\paragraph{Intervals}
In the main body of the music, all intervals must be consonant.
That is, we should only use 3rds, 5ths, 6ths, and 8ths.
This rule can be encoded as the \texttt{checkIntervals} predicate,
which again uses a new datatype \texttt{IntervalError} representing
the presence of a dissonant interval:

@@

\paragraph{Motion}
For every pair of two adjacent intervals, the cantus firmus and
counterpoint lines must not move in the same direction if the second
interval is a perfect one (i.e., unison, 5th, or 8th).
This is because such motion would draw too much attention to the
resulting interval.

@@

>>>>

In this section, we present an implementation of species counterpoint,
based on the formulation given by Fux~\citep{fux-cp}.
The idea is to represent ``good'' counterpoint  as a dependent record,
whose fields are the proofs that the counterpoint follows certain rules.
For space reason, we only describe two variants of species counterpoint;
other variants can be formalized in an analogous way.

@@

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Here is an example of first species counterpoint, where the lower line
stands for the cantus firmus and the upper line represents the
counterpoint\footnote{The cantus firmus is part of a German song
  called ``Froschgesang'' (Frog's song).}:

[Frog score]

We represent such music as a vector of pitch-interval pairs \texttt{(p , i)},
where \texttt{p} is a note in the cantus firmus, and \texttt{i} is the
distance between \texttt{p} and the corresponding counterpoint note.

\begin{alltt}
PitchInterval : Set
PitchInterval = Pitch \(\times\) Interval

frog-cp1 : Vector PitchInterval 7
frog-cp1 = (c 5 , per8) :: (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: (c 5 , per8) :: []
\end{alltt}

Note that the vector representation does not carry any correctness
guarantee of the music;
it requires extrinsic proofs stating that the music follows all the
rules for being good counterpoint.
Below, we discuss five such rules, showing what role they play
and how they can be represented using dependent types.

@@

\paragraph{Intervals}
In the main body of the music, we must ensure that vertical intervals
are harmonically pleasing.
all intervals must be consonant.
That is, we should only use 3rds, 5ths, 6ths, and 8ths.
This rule can be encoded as the \texttt{checkIntervals} predicate,
which again uses a new datatype \texttt{IntervalError} representing
the presence of a dissonant interval:

@@

\paragraph{Motion}

\begin{alltt}
data Motion : Set where
contrary : Motion
parallel : Motion
similar  : Motion
oblique  : Motion
\end{alltt}

For every pair of two adjacent intervals, the cantus firmus and
counterpoint lines must not move in the same direction if the second
interval is a perfect one (i.e., unison, 5th, or 8th).
This is because such motion would draw too much attention to the
resulting interval.

\begin{alltt}
data MotionError : Set where
  parallel : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) MotionError
  similar  : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) MotionError

motionCheck : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) Maybe MotionError
motionCheck i1 i2 with motion i1 i2 | isPerfect (proj\(\sb{\mathtt{2}}\) i2)
motionCheck i1 i2 | contrary | \_     = nothing
motionCheck i1 i2 | oblique  | \_     = nothing
motionCheck i1 i2 | parallel | false = nothing
motionCheck i1 i2 | parallel | true  = just (parallel i1 i2)
motionCheck i1 i2 | similar  | false = nothing
motionCheck i1 i2 | similar  | true  = just (similar i1 i2)

checkMotion : List PitchInterval → List MotionError
checkMotion = mapMaybe (uncurry motionCheck) ∘ pairs  
\end{alltt}

@@

\begin{alltt}
data UnisonError : Set where
  unison : Pitch \(\rightarrow\) UnisonError

unisonCheck : PitchInterval \(\rightarrow\) Maybe UnisonError
unisonCheck (p , i) = if (i == per1) then just (unison p) else nothing  
\end{alltt}
